---
title: "UCI HAR dataset cleaning Codebook"
author: "Biswajit"
date: "January 27, 2016"
output: html_document
---
##Introduction 

This file describes the data, the variables, and the work that has been performed to clean up the data

##Data Set Experiment

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data.

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain.

##For each record it provides:

- 

###The dataset includes the following files:

- CODEBOOK.rmd
- README.rmd
- run_analysis.R : R script to clean the provided data set
- tidydata_mean.txt : clean dataset as required.

###Variables

The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz.

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag).

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals).

These signals were used to estimate variables of the feature vector for each pattern:
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

### List of Variables 

[1] "subjectId"                                                           
 [2] "activityName"                                                        
 [3] "TimeDomain.Body.Acceleration.Mean.X"                                 
 [4] "TimeDomain.Body.Acceleration.Mean.Y"                                 
 [5] "TimeDomain.Body.Acceleration.Mean.Z"                                 
 [6] "TimeDomain.Body.Acceleration.StandardDeviation.X"                    
 [7] "TimeDomain.Body.Acceleration.StandardDeviation.Y"                    
 [8] "TimeDomain.Body.Acceleration.StandardDeviation.Z"                    
 [9] "TimeDomain.Gravity.Acceleration.Mean.X"                              
[10] "TimeDomain.Gravity.Acceleration.Mean.Y"                              
[11] "TimeDomain.Gravity.Acceleration.Mean.Z"                              
[12] "TimeDomain.Gravity.Acceleration.StandardDeviation.X"                 
[13] "TimeDomain.Gravity.Acceleration.StandardDeviation.Y"                 
[14] "TimeDomain.Gravity.Acceleration.StandardDeviation.Z"                 
[15] "TimeDomain.Body.AccelerationJerk.Mean.X"                             
[16] "TimeDomain.Body.AccelerationJerk.Mean.Y"                             
[17] "TimeDomain.Body.AccelerationJerk.Mean.Z"                             
[18] "TimeDomain.Body.AccelerationJerk.StandardDeviation.X"                
[19] "TimeDomain.Body.AccelerationJerk.StandardDeviation.Y"                
[20] "TimeDomain.Body.AccelerationJerk.StandardDeviation.Z"                
[21] "TimeDomain.BodyGyro.Mean.X"                                          
[22] "TimeDomain.BodyGyro.Mean.Y"                                          
[23] "TimeDomain.BodyGyro.Mean.Z"                                          
[24] "TimeDomain.BodyGyro.StandardDeviation.X"                             
[25] "TimeDomain.BodyGyro.StandardDeviation.Y"                             
[26] "TimeDomain.BodyGyro.StandardDeviation.Z"                             
[27] "TimeDomain.BodyGyroJerk.Mean.X"                                      
[28] "TimeDomain.BodyGyroJerk.Mean.Y"                                      
[29] "TimeDomain.BodyGyroJerk.Mean.Z"                                      
[30] "TimeDomain.BodyGyroJerk.StandardDeviation.X"                         
[31] "TimeDomain.BodyGyroJerk.StandardDeviation.Y"                         
[32] "TimeDomain.BodyGyroJerk.StandardDeviation.Z"                         
[33] "TimeDomain.Body.AccelerationMagnitude.Mean"                          
[34] "TimeDomain.Body.AccelerationMagnitude.StandardDeviation"             
[35] "TimeDomain.Gravity.AccelerationMagnitude.Mean"                       
[36] "TimeDomain.Gravity.AccelerationMagnitude.StandardDeviation"          
[37] "TimeDomain.Body.AccelerationJerkMagnitude.Mean"                      
[38] "TimeDomain.Body.AccelerationJerkMagnitude.StandardDeviation"         
[39] "TimeDomain.BodyGyroMagnitude.Mean"                                   
[40] "TimeDomain.BodyGyroMagnitude.StandardDeviation"                      
[41] "TimeDomain.BodyGyroJerkMagnitude.Mean"                               
[42] "TimeDomain.BodyGyroJerkMagnitude.StandardDeviation"                  
[43] "FrequencyDomain.Body.Acceleration.Mean.X"                            
[44] "FrequencyDomain.Body.Acceleration.Mean.Y"                            
[45] "FrequencyDomain.Body.Acceleration.Mean.Z"                            
[46] "FrequencyDomain.Body.Acceleration.StandardDeviation.X"               
[47] "FrequencyDomain.Body.Acceleration.StandardDeviation.Y"               
[48] "FrequencyDomain.Body.Acceleration.StandardDeviation.Z"               
[49] "FrequencyDomain.Body.Acceleration.MeanFrequency.X"                   
[50] "FrequencyDomain.Body.Acceleration.MeanFrequency.Y"                   
[51] "FrequencyDomain.Body.Acceleration.MeanFrequency.Z"                   
[52] "FrequencyDomain.Body.AccelerationJerk.Mean.X"                        
[53] "FrequencyDomain.Body.AccelerationJerk.Mean.Y"                        
[54] "FrequencyDomain.Body.AccelerationJerk.Mean.Z"                        
[55] "FrequencyDomain.Body.AccelerationJerk.StandardDeviation.X"           
[56] "FrequencyDomain.Body.AccelerationJerk.StandardDeviation.Y"           
[57] "FrequencyDomain.Body.AccelerationJerk.StandardDeviation.Z"           
[58] "FrequencyDomain.Body.AccelerationJerk.MeanFrequency.X"               
[59] "FrequencyDomain.Body.AccelerationJerk.MeanFrequency.Y"               
[60] "FrequencyDomain.Body.AccelerationJerk.MeanFrequency.Z"               
[61] "FrequencyDomain.BodyGyro.Mean.X"                                     
[62] "FrequencyDomain.BodyGyro.Mean.Y"                                     
[63] "FrequencyDomain.BodyGyro.Mean.Z"                                     
[64] "FrequencyDomain.BodyGyro.StandardDeviation.X"                        
[65] "FrequencyDomain.BodyGyro.StandardDeviation.Y"                        
[66] "FrequencyDomain.BodyGyro.StandardDeviation.Z"                        
[67] "FrequencyDomain.BodyGyro.MeanFrequency.X"                            
[68] "FrequencyDomain.BodyGyro.MeanFrequency.Y"                            
[69] "FrequencyDomain.BodyGyro.MeanFrequency.Z"                            
[70] "FrequencyDomain.Body.AccelerationMagnitude.Mean"                     
[71] "FrequencyDomain.Body.AccelerationMagnitude.StandardDeviation"        
[72] "FrequencyDomain.Body.AccelerationMagnitude.MeanFrequency"            
[73] "FrequencyDomain.BodyBody.AccelerationJerkMagnitude.Mean"             
[74] "FrequencyDomain.BodyBody.AccelerationJerkMagnitude.StandardDeviation"
[75] "FrequencyDomain.BodyBody.AccelerationJerkMagnitude.MeanFrequency"    
[76] "FrequencyDomain.BodyBodyGyroMagnitude.Mean"                          
[77] "FrequencyDomain.BodyBodyGyroMagnitude.StandardDeviation"             
[78] "FrequencyDomain.BodyBodyGyroMagnitude.MeanFrequency"                 
[79] "FrequencyDomain.BodyBodyGyroJerkMagnitude.Mean"                      
[80] "FrequencyDomain.BodyBodyGyroJerkMagnitude.StandardDeviation"         
[81] "FrequencyDomain.BodyBodyGyroJerkMagnitude.MeanFrequency"

## Code for run_analysis.R

###steps to perform 

1.   Merges the training and the test sets to create one data set.
2.   Extracts only the measurements on the mean and standard deviation for each measurement.
3.   Uses descriptive activity names to name the activities in the data set
4.   Appropriately labels the data set with descriptive variable names.
5.   From the data set in step 4, creates a second, independent tidy data set with the average of each variable       for each activity and each subject.

###1. extract the data and merge it.

```{r}
#load required libraries
library(plyr)

# reading all the data to the workspace
activityLabels <- read.table('activity_labels.txt',header = FALSE) # read activity names
names(activityLabels)<- c("activityId", "activityName")


features <- read.table('features.txt', header = FALSE) #read the feature list
names(features) <- c("featureId","featureName")

#read train data
trainX <- read.table('train/X_train.txt', header = FALSE)
trainY <- read.table('train/y_train.txt')
trainSubject <- read.table("./train/subject_train.txt", header = FALSE)

# assign required names to col names
names(trainSubject) <- c("subjectId") 
names(trainY) <- c("activityId")
names(trainX) <- features[,2]

# make the combined train data
trainData <- cbind(trainY,trainSubject,trainX)

#read test data
testX <- read.table('test/X_test.txt', header = FALSE)
testY <- read.table('test/y_test.txt')
testSubject <- read.table("./test/subject_test.txt", header = FALSE)

#assign required names to cols
names(testSubject) <- c("subjectId")
names(testY) <- c("activityId")
names(testX) <- features[,2]

# make the combined test data
testData <- cbind(testY,testSubject,testX) 

#merge test and train data
mergedData <- rbind(trainData, testData)
```

###2.   Extracts only the measurements on the mean and standard deviation for each measurement.

```{r}
# extract the mean and std cols from merge data
mergeData_mean_std <- mergedData[,grepl("mean|std|subjectId|activityId",names(mergedData))]
```

###3.   Uses descriptive activity names to name the activities in the data set

```{r}
#use the descriptive activity names in the merged data set
mergeDataMeanStd_descriptive <- join(activityLabels,mergeData_mean_std, by = "activityId")
mergeDataMeanStd_descriptive <- mergeDataMeanStd_descriptive[,-1]
```


###4.   Appropriately labels the data set with descriptive variable names.

```{r}
#clean the column names 
names(mergeDataMeanStd_descriptive) <- gsub("\\(|\\)","",names(mergeDataMeanStd_descriptive),perl = TRUE)
names(mergeDataMeanStd_descriptive) <- make.names(names(mergeDataMeanStd_descriptive))

names(mergeDataMeanStd_descriptive) <- gsub('Acc',".Acceleration",names(mergeDataMeanStd_descriptive))
names(mergeDataMeanStd_descriptive) <- gsub('Mag',"Magnitude",names(mergeDataMeanStd_descriptive))
names(mergeDataMeanStd_descriptive) <- gsub('^t',"TimeDomain.",names(mergeDataMeanStd_descriptive))
names(mergeDataMeanStd_descriptive) <- gsub('^f',"FrequencyDomain.",names(mergeDataMeanStd_descriptive))
names(mergeDataMeanStd_descriptive) <- gsub('\\.mean',".Mean",names(mergeDataMeanStd_descriptive))
names(mergeDataMeanStd_descriptive) <- gsub('\\.std',".StandardDeviation",names(mergeDataMeanStd_descriptive))
names(mergeDataMeanStd_descriptive) <- gsub('Freq\\.',"Frequency.",names(mergeDataMeanStd_descriptive))
names(mergeDataMeanStd_descriptive) <- gsub('Freq$',"Frequency",names(mergeDataMeanStd_descriptive))

```

###5.   From the data set in step 4, creates a second, independent tidy data set with the average of each               variable for each activity and each subject.

```{r}
# create a new copy of the final merged data 
mergeDataMeanStd_descriptive2 <- mergeDataMeanStd_descriptive

#creates a second, independent tidy data set with the average of each variable for each 
#activity and each subject.
tidy<- ddply(mergeDataMeanStd_descriptive2,c("subjectId","activityName"),numcolwise(mean))

#save it to a file
write.table(tidy, file = "tidydata_mean.txt")
```
